RegressionPvNP

### 2017-05-10 RegressionPvNP.Rmd
Using deviance measure rather than performance to select model

```{r setup, include=FALSE}
library(ggplot2)
library(ggfortify)
library(plyr)
library(pander)
#library(Hmisc)
library(reshape2)
library(gridExtra)
library(GenomicRanges)
library(glmnet)

@@

sum.patient.data = summarise.patient.info(patient.info)
sum.patient.data = as.data.frame(subset(sum.patient.data, Patient %in% names(patient.data))) ## For now
pander(sum.patient.data[,c('Patient', 'Status', 'start.year', 'end.year','total.samples','total.endos','highest.path')])

# Missing some of the samples as they aren't all sequenced yet
for (pt in names(patient.data)) {
  patient.data[[pt]]$info = patient.data[[pt]]$info[patient.data[[pt]]$info$Samplename %in% colnames(patient.data[[pt]]$seg.vals)[-(1:5)],]
}
```

```{r apclust-func, echo=F, message=F, warning=F}
#suppressPackageStartupMessages( library(apcluster) )

apclust.data<-function(segdata, samples) {
  x1 = segdata[,intersect(colnames(segdata), samples)]
  x1 = x1[, samples]
  rownames(x1) = (segdata[,c(1:4)] %>%
    rowwise() %>%
    mutate(location=paste(paste(chrom, arm, sep=''), '.', start.pos, '-', end.pos, sep='')))$location
  return(x1)  
}

# apclust<-function(segdata) {
#   x1 = apclust.data(segdata)
#   # q=0 minimize off-diagonal similarity
#   ac = apcluster(negDistMat(r=2), x1, details=T, convits=25, q=0)
#   return(list('apres'=ac, 'data'=x1))
# }

@@

tile.sample.segments<-function(tileFile, sample.data) {
  if (file.exists(tileFile)) {
    load(tileFile, verbose=T)
  } else {
    grList = lapply(sample.data, function(df) {
      df$info = arrange(df$info, Endoscopy.Year, Pathology)

@@

Data matrix currently:
`r pander(dysplasia.df[1:10, 1:5], caption=paste("dimensions:", paste(dim(dysplasia.df), collapse=', ')), justify='left')`

@@

plot.multi.cv.glmnet<-function(x,y,alpha=c(0,1), family="binomial", title='cv.glmnet') {
  plots <- list(  )
  cv.models = data.frame(matrix(ncol=3,nrow=0,dimnames=list(c(), c('class','lambda','cvm'))))
  for (a in alpha) {
    cv = cv.glmnet(x, y, alpha=a, family=family, standardize=F)
    cv.models = rbind(cv.models, cbind('class'=paste('cv',a,sep=''), 'lambda'=log(cv$lambda), 'cvm'=cv$cvm))
    plots[[as.character(a)]] = autoplot(cv, main=paste('alpha',a), ylab=cv$name)
  }
  cv.models[c('lambda','cvm')] = lapply(cv.models[c('lambda','cvm')], function(x) as.numeric(as.character(x)))

@@

gp = plot.multi.cv.glmnet(dysplasia.df, labels, alpha=c(0,0.5,0.8,1), title='cv.glmnet P vs NP')

@@

@@

crossvalidate.by.patient<-function(x,y,lambda,pts,a=1,nfolds=10, splits=5, fit=NULL, minR=0.2) {
  message(paste("Running", splits, "splits",nfolds,"times on", paste(dim(x), collapse=':'), 'alpha=',a ))
  fit.e = list()
  pts = create.patient.sets(pts, nfolds, splits, minR)
  cv.pred = (matrix(nrow=0, ncol=length(lambda)))
  for (n in 1:nfolds) {  
    message(paste(n, "fold"))
    setCol = grep(paste('^fold.',n,'$',sep=''), colnames(pts))

    cv.class = matrix(nrow=splits, ncol=length(lambda))
    for (i in 1:splits) { 
      message(paste(i, "split"))
      test.rows = which(rownames(x) %in% pts[which(pts[,setCol] == i), 'Samplename'])
      test = x[test.rows,]
      training = x[-test.rows,]
      # pre-spec lambda seq
      fit <- glmnet(training, y[-test.rows], lambda=lambda, family='binomial', alpha=a) 
      # autoplot(fit) + theme(legend.position="none")
      # careful: matrix
      pred <- pi.hat(predict(fit, test, type='link')) 
      cv.class[i,] = apply(pred, 2, function(p) {
        # Confusion matrix: quantitiative, pos results + neg results / number of test rows
        (p%*%labels[test.rows] + (1-p) %*% (1-labels[test.rows]))/ length(test.rows)
      })

      fit.e[[length(fit.e)+1]] = fit
    }
  cv.pred = rbind(cv.pred, cv.class)    
  }

  df = cbind.data.frame('lambda.at'=1:ncol(cv.pred),
                        'mean'= colMeans(cv.pred), 
                        'sme'= apply(cv.pred, 2, sd)/sqrt(nrow(cv.pred)), 
                        'sd'= apply(cv.pred, 2, sd), 
                        'lambda'= lambda)

  df2 = df
  df2$lambda = log(lambda)
  df2$mean = 1-df2$mean

  df2$lambda.min = df2$mean == min(df2$mean)
  df2$lambda.1se = FALSE
  df2[df2$lambda < subset(df2, mean == min(mean))$lambda-sd(df2$lambda),][1,'lambda.1se'] = TRUE

  lambda.min = lambda[subset(df2, lambda.min == T)$lambda.at]
  lambda.1se = lambda[subset(df2, lambda.1se == T)$lambda.at]
  nzcf = lapply(fit.e, non.zero.coef, s=lambda.min)

  gp = ggplot(df2, aes(y=mean,x=lambda)) + geom_point() + 
    geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme)) + 
    geom_point(data=subset(df2, lambda.min == T), aes(y=mean, x=lambda, colour="lambda.min"), size=2 ) +          
    geom_vline(xintercept = subset(df2, lambda.min == T)$lambda, colour='grey') +
    annotate("text", x=subset(df2, lambda.min == T)$lambda, y=0.35, 
             label=round(df[subset(df2, lambda.min == T)$lambda.at,'mean'],3))

  if (length(lambda.1se) > 0) {
    gp = gp + geom_point(data=subset(df2, lambda.1se == T), aes(y=mean, x=lambda, colour='lambda.1se'), size=2 ) +
      geom_vline(xintercept = subset(df2, lambda.1se == T)$lambda, colour='grey') +
      annotate("text", x=subset(df2, lambda.1se == T)$lambda, y=0.35, 
               label=round(df[subset(df2, lambda.1se == T)$lambda.at,'mean'],3)) 
  }
   gp = gp + theme( legend.position='bottom') + 
          scale_colour_discrete(name = "") +
          labs(title=paste(splits,' splits, ',nfolds,' folds, alpha=',a, sep=''), y='1-mean (P)', x='log(lambda)')

  if (!is.null(fit)) {
    df2$nzcoef = sapply(df$lambda, function(l) length(non.zero.coef(fit, l)))

    coef.min = subset(df2, lambda.min == T)$nzcoef
    coef.1se = subset(df2, lambda.1se == T)$nzcoef
    d = coef.1se-coef.min

    coef.text = arrange(subset(df2,nzcoef %in% c(0,  coef.min, coef.1se, max(nzcoef) )), -lambda.min, -lambda.1se)
    coef.text = arrange(coef.text[!duplicated(coef.text$nzcoef),], -nzcoef)

    gp = gp + annotate("text", y=max(df2$mean), x=coef.text$lambda, label=coef.text$nzcoef)
  }

  return(list('max.cm'=df[which.max(df$mean)-1,'mean'], 
              'lambda.min'=lambda.min, 'lambda.1se'=lambda.1se, 'lambdas'=df, 'plot'=gp, 'non.zero.cf'=nzcf))
}

coef.stability<-function(opt, nz.list) {
  for (i in 1:length(nz.list)) {
    df = as.data.frame(nz.list[[i]])
    colnames(df) = i
    opt = merge(opt, df, by='row.names', all.x=T)
    rownames(opt) = opt$Row.names
    opt$Row.names = NULL
    opt[,(i+1)] = as.integer(!is.na(opt[,(i+1)]))
  }
  opt = opt[order(sapply(rownames(opt), function(x) as.numeric(unlist(strsplit(x, ':'))[1])  )),]
  return(opt)
}
```

```{r xvalpt, message=F, warning=F, echo=T, fig.height=12, fig.width=12}
plots = list()
coefs = list()
folds = 10; splits=5
alpha.values = c(0,0.5,0.7,0.8,0.9,1)
for (a in alpha.values) {
  fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=1000, family='binomial') # all patients
  autoplot(fit0, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none') 
  lambda0 <- fit0$lambda

  cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=lambda0, pts=pts, a=a, nfolds=folds, splits=splits, fit=fit0)

  coef.1se = as.data.frame(non.zero.coef(fit0, cv.patient$lambda.1se))
  coefs[[as.character(a)]] = coef.stability(coef.1se, cv.patient$non.zero.cf)

  plots[[as.character(a)]] = cv.patient$plot + theme(legend.position='bottom') + ggtitle(paste('alpha=',a,sep=''))
}

do.call(grid.arrange, c(plots, top='All samples, 10fold, 5 splits', ncol=2))

@@

```{r noHGD, echo=T, message=F, warning=F, fig.height=8, fig.width=8}
`%nin%` <- Negate(`%in%`)

@@

  plots[[as.character(a)]] = cv.nohgd$plot + ggtitle(paste('alpha=',a,sep=''))
  coef.1se = as.data.frame(non.zero.coef(fit1, cv.nohgd$lambda.1se))
  coefs[[as.character(a)]] = coef.stability(coef.1se, cv.nohgd$non.zero.cf)
}

@@

@@

plots = list()
coefs = list()
for (a in alpha.values) {
  fit2 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
  autoplot(fit2, xvar='lambda', main='fit1, no HGD/IMC samples') + theme(legend.position='none') 
  lambda2 <- fit2$lambda

@@

  plots[[as.character(a)]] = cv.nolgd$plot + ggtitle(paste('alpha=',a,sep=''))

@@

@@

### Leave one patient out

Using the 'all sample' fit, run a LOO (patient)

```{r loop, echo=T, message=F, warning=F, fig.height=6, fig.width=6, eval=F}
fit0 <- glmnet(dysplasia.df, labels, alpha=1, nlambda=200, family='binomial') # all patients

# Randomize patients
pt.samp = sample(sum.patient.data$Patient)
performance.at.1se = c()
coefs = list()
for (pt in pt.samp) {
  samples = as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Patient != pt)$Patient], function(pt) pt$info$Samplename)))

  cv.loo = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fit0$lambda, pts= subset(pts, Samplename %in% samples), a=1, nfolds=20, splits=5, fit=fit0)

  performance.at.1se = c(performance.at.1se, subset(cv.loo$lambdas, lambda == cv.loo$lambda.1se)$mean)

  coef.1se = as.data.frame(non.zero.coef(fit0, cv.loo$lambda.1se))
  coefs[[as.character(a)]] = coef.stability(coef.1se, cv.loo$non.zero.cf)
}

ggplot(as.data.frame(performance.at.1se), aes(y=performance.at.1se, x='LOO')) + ylim(0.6,0.8) +
  geom_boxplot( fill='lightblue', color='darkgrey', outlier.fill=NA, outlier.color=NA) + geom_jitter() +
    labs(y='performance at lambda.1se', x='', title='Performance for Leave One (patient) Out, at alpha=1')

@@

fit0 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=1, nlambda=200, family='binomial') # all patients
autoplot(fit0) + theme(legend.position = 'none')

cv.back.few = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fit0$lambda, pts=subset(pts, Samplename %in% samples), splits=5, nfolds=20, a=1, fit=fit0)

# Many
samples = c(nonps, as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'P' & total.samples >= m)$Patient], function(pt) pt$info$Samplename))))
cv.back.many = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda0, subset(patients, Samplename %in% samples))

grid.arrange(cv.back.few$plot + ggtitle(paste(nrow(subset(sum.patient.data, Status == 'P' & total.samples < m))," Progressors with < median # samples")), 
             cv.back.many$plot + ggtitle(paste(nrow(subset(sum.patient.data, Status == 'P' & total.samples >= m)), "Progressors with >= median # samples")), top=paste("Median number of samples:", m))

@@

## Peel back timepoints

So here, do I use the initial fit (all samples), or should I be using the noHGD fit...

```{r timepoints, echo=T, message=F, warning=F, eval=F}

get.next.samples<-function(patient.data, i) {
  as.vector(unlist(sapply(patient.data, function(pd) {
    j = i
    pd$info = arrange(pd$info, Endoscopy.Year, Pathology)

    if (j >= nrow(pd$info)) j = nrow(pd$info)

    pd$info[1:(nrow(pd$info)-j), 'Samplename']
  }) ))
}

nonps = get.next.samples(patient.data[subset(sum.patient.data, Status == 'NP')$Patient], 0)
fpr = data.frame(matrix(ncol=3,nrow=0))
for (i in c(0,1,seq(2,35,3))) {
  ps = get.next.samples(patient.data[subset(sum.patient.data, Status == 'P')$Patient], i)
  #samples = get.next.samples(patient.data, i)
  samples = c(nonps, ps)
  cv.back = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda0, subset(patients, Samplename %in% samples))
  fpr = rbind(fpr, c(cv.back$max.cm, sum(cv.back$lambdas[ which.max(cv.back$lambdas$mean)-1, c('sme')]), length(ps)))
#  plots[[ as.character(i) ]] = cv.back$plot + ggtitle(paste(length(ps), "progressor samples. mean f1:", cv.back$f1))
}
colnames(fpr) = c('performance','error','n.prog.samples')

grid.arrange(ggplot(fpr, aes(performance,n.prog.samples)) + geom_point(),
             ggplot(fpr, aes(performance, error)) + geom_point(), top='Peel back samples')

pander(fpr, justify='left', caption="Removing timepoints from the model and retraining")
```

Can look then at a single point that may be an example of the performance of the model.

```{r singletimepoint, echo=T, message=F, warning=F, eval=F}
ps = get.next.samples(patient.data[subset(sum.patient.data, Status == 'P')$Patient], 11)
samples = c(nonps, ps)
cv.back = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda0, subset(patients, Samplename %in% samples))

cv.back$plot + ggtitle(paste("CV Performance - ", length(ps), "progressor samples"))

info = subset(patient.info, Samplename %in% samples)

coef.opt.stability = coef.stability(coef.opt, cv.back$non.zero.cf)

>>>>

```{r setup, include=FALSE}
library(ggplot2)
library(ggfortify)
library(plyr)
library(pander)
library(reshape2)
library(gridExtra)
library(GenomicRanges)
library(glmnet)

@@

sum.patient.data = summarise.patient.info(patient.info)
sum.patient.data = as.data.frame(subset(sum.patient.data, Patient %in% names(patient.data))) ## For now

pander(sum.patient.data[,c('Patient', 'Status', 'start.year', 'end.year','total.samples','total.endos','highest.path')], justify='left')

# Missing some of the samples as they aren't all sequenced yet
for (pt in names(patient.data)) {
  #print(pt)
  #print(patient.data[[pt]]$info$Samplename %in% colnames(patient.data[[pt]]$seg.vals)[-(1:5)])
  patient.data[[pt]]$info = patient.data[[pt]]$info[patient.data[[pt]]$info$Samplename %in% colnames(patient.data[[pt]]$seg.vals)[-(1:5)],]
}
```

```{r apclust-func, echo=F, message=F, warning=F}
apclust.data<-function(segdata, samples) {
  x1 = segdata[,intersect(colnames(segdata), samples)]
  x1 = x1[, samples]
  rownames(x1) = (segdata[,c(1:4)] %>%
                    rowwise() %>%
                    mutate(location=paste(paste(chrom, arm, sep=''), '.', start.pos, '-', end.pos, sep='')))$location
  return(x1)  
}
```

@@

tile.sample.segments<-function(tileFile, sample.data) {
  if (file.exists(tileFile)) {
    message(paste("Loading", tileFile))
    load(tileFile, verbose=T)
  } else {
    message("Tiling patient data...")
    grList = lapply(sample.data, function(df) {
      df$info = arrange(df$info, Endoscopy.Year, Pathology)

@@

Example subset of the data matrix:
`r pander(dysplasia.df[1:10, 1:5], caption=paste("dimensions:", paste(dim(dysplasia.df), collapse=', ')), justify='left')`

@@

plot.multi.cv.glmnet<-function(x,y,alpha=c(0,1), family="binomial", title='cv.glmnet') {
  plots <- list(  )
  cv.models = data.frame(matrix(ncol=3,nrow=0,dimnames=list(c(), c('class','lambda','cvm'))))
  for (a in alpha) {
    cv = cv.glmnet(x, y, alpha=a, family=family)
    cv.models = rbind(cv.models, cbind('class'=paste('cv',a,sep=''), 'lambda'=log(cv$lambda), 'cvm'=cv$cvm))
    plots[[as.character(a)]] = autoplot(cv, main=paste('alpha',a), ylab=cv$name)
  }
  cv.models[c('lambda','cvm')] = lapply(cv.models[c('lambda','cvm')], function(x) as.numeric(as.character(x)))

@@

gp = plot.multi.cv.glmnet(x=dysplasia.df, y=labels, alpha=c(0,0.5,0.8,1), title='cv.glmnet P vs NP')

@@

coef.stability<-function(opt, nz.list) {
  for (i in 1:length(nz.list)) {
    df = as.data.frame(nz.list[[i]])
    colnames(df) = i
    opt = merge(opt, df, by='row.names', all.x=T)
    rownames(opt) = opt$Row.names
    opt$Row.names = NULL
    opt[,(i+1)] = as.integer(!is.na(opt[,(i+1)]))
  }
  opt = opt[order(sapply(rownames(opt), function(x) as.numeric(unlist(strsplit(x, ':'))[1])  )),]
  return(opt)
}

@@

crossvalidate.by.patient<-function(x,y,lambda,pts,a=1,nfolds=10, splits=5, fit=NULL, minR=0.2, select='performance') {
  if (nfolds > 5) minR = 0.1
  message(paste("Running", splits, "splits",nfolds,"times on", paste(dim(x), collapse=':'), 'alpha=',a ))
  fit.e = list()
  tpts = create.patient.sets(pts, nfolds, splits, minR)
  cv.pred = (matrix(nrow=0, ncol=length(lambda)))
  cv.binomial.deviance = (matrix(nrow=0, ncol=length(lambda)))
  for (n in 1:nfolds) {  
    message(paste(n, "fold"))
    setCol = grep(paste('^fold.',n,'$',sep=''), colnames(tpts))

    cv.class = matrix(nrow=splits, ncol=length(lambda))
    deviance = matrix(nrow=splits, ncol=length(lambda))
    for (i in 1:splits) { 
      message(paste(i, "split"))
      test.rows = which(rownames(x) %in% tpts[which(tpts[,setCol] == i), 'Samplename'])
      test = x[test.rows,]
      training = x[-test.rows,]
      # pre-spec lambda seq
      fitCV <- glmnet(training, y[-test.rows], lambda=lambda, family='binomial', alpha=a) 
      # autoplot(fit) + theme(legend.position="none")

      # Confusion matrix: quantitiative, pos results + neg results / number of test rows
      pred <- pi.hat(predict(fitCV, test, type='link')) 
      cv.class[i,] = apply(pred, 2, function(p) {
        (p%*%y[test.rows] + (1-p) %*% (1-y[test.rows]))/length(test.rows)
      })

      # Binomial deviance, lifted from cv.lognet
      predmat = predict(fitCV, test, type='response')
      prob_min = 1e-05; prob_max = 1 - prob_min

      predmat = pmin(pmax(predmat, prob_min), prob_max)

      yT = as.factor(y[test.rows])
      dev = apply(predmat, 2, function(p)  -2*((yT==1)*log(p)+(yT==0)*log(1-p)) )

      deviance[i,] = apply(dev, 2, weighted.mean, w=rep(1, nrow(dev)), na.rm=T)

      fit.e[[length(fit.e)+1]] = fit
    }
    cv.pred = rbind(cv.pred, cv.class)    
    cv.binomial.deviance = rbind(cv.binomial.deviance, deviance)
  }

  # Not really sure what to do with the binomial deviance now...
  df = cbind.data.frame('lambda.at'=1:ncol(cv.pred),
                        'mean'= colMeans(cv.pred), 
                        'sme'= apply(cv.pred, 2, sd)/sqrt(nrow(cv.pred)), 
                        'sd'= apply(cv.pred, 2, sd), 
                        'lambda'= lambda,
                        'log.lambda' = log(lambda),
                        'mean.b.dev' = colMeans(cv.binomial.deviance),
                        'sd.b.dev' = apply(cv.binomial.deviance, 2,sd),
                        'sme.b.dev' = apply(cv.binomial.deviance, 2, sd)/sqrt(nrow(cv.binomial.deviance)),
                        'lambda.min' = F, 'lambda.1se' = F)
  #df$lambda.1se = FALSE
  #df$`lambda-1se` = FALSE

  if (select == 'performance') {
    df$lambda.min = df$mean == max(df$mean)
    df[df$log.lambda < subset(df, mean == max(mean))$log.lambda+sd(df$log.lambda),][1, 'lambda.1se'] = TRUE
    #df[df$log.lambda < subset(df, mean == max(mean))$log.lambda-sd(df$log.lambda),][1, 'lambda-1se'] = TRUE
  } else if (select == 'deviance') {
    df$lambda.min = df$mean.b.dev == min(df$mean.b.dev)
    df[df$mean.b.dev >= df[df$lambda.min == T, 'mean.b.dev'] + sd(df$mean.b.dev), ][1, 'lambda.1se'] = T
  }

  lambda.min = lambda[subset(df, lambda.min == T)$lambda.at]
  lambda.1se = lambda[subset(df, lambda.1se == T)$lambda.at]

  nzcf = lapply(fit.e, non.zero.coef, s=lambda.1se)

  plots = plot.patient.cv(df, fit)
  plots$performance = plots$performance + theme(legend.position='bottom') + scale_colour_discrete(name = "") +
    labs(title=paste(splits,' splits, ',nfolds,' folds, alpha=',a, sep=''), y='mean pred.', x='log(lambda)') 
  plots$deviance = plots$deviance + theme(legend.position='bottom') + scale_colour_discrete(name="") + 
    labs(title=paste(splits,' splits, ',nfolds,' folds, alpha=',a, sep=''), y='mean Binomial Deviance', x='log(lambda)')

  return(list('max.cm'=df[which.max(df$mean)-1,'mean'], 
              'lambda.min'=lambda.min, 'lambda.1se'=lambda.1se, 'lambdas'=df, 'plot'=plots$performance, 'deviance.plot'=plots$deviance, 'non.zero.cf'=nzcf))
}

plot.patient.cv<-function(df, fit=NULL) {
    gp = ggplot(df, aes(y=mean,x=log.lambda)) + geom_point() + geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme)) + 
    geom_point(data=subset(df, lambda.min == T), aes(y=mean, x=log.lambda, colour="lambda.min"), size=2 ) +  
    geom_vline(xintercept = subset(df, lambda.min == T)$log.lambda, colour='grey') +
    annotate("text", x=subset(df, lambda.min == T)$log.lambda, y=min(df$mean)+sd(df$mean), 
             label=round(df[subset(df, lambda.min == T)$lambda.at,'mean'],3))

    gpD = ggplot(df, aes(y=mean.b.dev,x=log.lambda)) + geom_point() +
    geom_errorbar(aes(ymin=mean.b.dev-sme.b.dev, ymax=mean.b.dev+sme.b.dev)) + 
    geom_point(data=subset(df, lambda.min == T), aes(y=mean.b.dev, x=log.lambda, colour="lambda.min"), size=2 ) +          
    geom_vline(xintercept = subset(df, lambda.min == T)$log.lambda, colour='grey') +
    annotate("text", x=subset(df, lambda.min == T)$log.lambda, y=min(df$mean)+sd(df$mean), 
             label=round(df[subset(df, lambda.min == T)$lambda.at,'mean'],3))

  if (nrow(subset(df, lambda.1se == T)) > 0) {
    gp = gp + geom_point(data=subset(df, lambda.1se == T), aes(y=mean, x=log.lambda, colour="lambda.min+1se"), size=2 ) +
      geom_vline(xintercept = subset(df, lambda.1se == T)$log.lambda, colour='grey') +
      annotate("text", x=subset(df, lambda.1se == T)$log.lambda, y=min(df$mean)+sd(df$mean), 
               label=round(df[subset(df, lambda.1se == T)$lambda.at,'mean'],3))

    gpD = gpD + geom_point(data=subset(df, lambda.1se == T), aes(y=mean.b.dev, x=log.lambda, colour="lambda.min+1se"), size=2 ) +
      geom_vline(xintercept = subset(df, lambda.1se == T)$log.lambda, colour='grey') +
      annotate("text", x=subset(df, lambda.1se == T)$log.lambda, y=min(df$mean)+sd(df$mean), 
               label=round(df[subset(df, lambda.1se == T)$lambda.at,'mean'],3)) 
  }

  if (nrow(subset(df, `lambda-1se` == T)) > 0) {
      gp = gp + geom_point(data=subset(df, `lambda-1se` == T), aes(y=mean, x=log.lambda, colour="lambda.min-1se"), size=2 ) +
      geom_vline(xintercept = subset(df, `lambda-1se` == T)$log.lambda, colour='grey') +
      annotate("text", x=subset(df, `lambda-1se` == T)$log.lambda, y=min(df$mean)+sd(df$mean), 
               label=round(df[subset(df, `lambda-1se` == T)$lambda.at,'mean'],3))

      gpD = gpD + geom_point(data=subset(df, `lambda-1se` == T), aes(y=mean.b.dev, x=log.lambda, colour="lambda.min-1se"), size=2 ) +
      geom_vline(xintercept = subset(df, `lambda-1se` == T)$log.lambda, colour='grey') +
      annotate("text", x=subset(df, `lambda-1se` == T)$log.lambda, y=min(df$mean)+sd(df$mean), 
               label=round(df[subset(df, `lambda-1se` == T)$lambda.at,'mean'],3)) 
  }

  gp = gp + theme(legend.position='bottom') + scale_colour_discrete(name = "") 
  gpD = gpD + theme(legend.position='bottom') + scale_colour_discrete(name="")

  if (!is.null(fit)) {
    df$nzcoef = sapply(df$lambda, function(l) length(non.zero.coef(fit, l)))

    coef.min = subset(df, lambda.min == T)$nzcoef
    coef.1se = subset(df, lambda.1se == T)$nzcoef
    coef.m.1se = subset(df, `lambda-1se` == T)$nzcoef

    d = coef.1se-coef.min

    coef.text = arrange(subset(df,nzcoef %in% c(0,  coef.min, coef.1se, coef.m.1se, max(nzcoef) )), -lambda.min, -lambda.1se)
    coef.text = arrange(coef.text[!duplicated(coef.text$nzcoef),], -nzcoef)

    gp = gp + annotate("text", y=min(df$mean), x=coef.text$log.lambda, label=coef.text$nzcoef, color='darkblue')
    gpD = gpD + annotate("text", y=max(df$mean.b.dev), x=coef.text$log.lambda, label=coef.text$nzcoef, color='darkblue')
  }
return(list('performance'=gp, 'deviance'=gpD))
}

```{r xvalpt, message=F, warning=F, echo=T, fig.height=12, fig.width=12}
performance.plots = list(); coefs = list(); deviance.plots = list()
plots = list()
folds = 10; splits=5
alpha.values = c(0,0.5,0.7,0.8,0.9,1)
for (a in alpha.values) {
  fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=1000, family='binomial') # all patients
  autoplot(fit0, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none')

  cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=fit0$lambda, pts=pts, a=a, nfolds=folds, splits=splits, fit=fit0, select='deviance')

  lambda.opt = ifelse( length(cv.patient$lambda.1se) > 0, cv.patient$lambda.1se, cv.patient$lambda.min )

  coef.opt = as.data.frame(non.zero.coef(fit0, lambda.opt))
  coefs[[as.character(a)]] = coef.stability(coef.opt, cv.patient$non.zero.cf)

  #performance.plots[[as.character(a)]] = cv.patient$plot + ggtitle(paste('alpha=',a,sep=''))
  #deviance.plots[[as.character(a)]] = cv.patient$deviance.plot + ggtitle(paste('alpha=',a,sep=''))

  plots[[as.character(a)]] = arrangeGrob(cv.patient$plot+ggtitle('CV Performance'), cv.patient$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)
}

do.call(grid.arrange, c(plots, top='All samples, 10fold, 5 splits'))

@@

```{r noHGD, echo=T, message=F, warning=F, fig.height=12, fig.width=12}
`%nin%` <- Negate(`%in%`)

@@

  #plots[[as.character(a)]] = cv.nohgd$plot + ggtitle(paste('alpha=',a,sep=''))

  plots[[as.character(a)]] = arrangeGrob(cv.patient$plot+ggtitle('CV Performance'), cv.patient$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)

  coef.1se = as.data.frame(non.zero.coef(fit1, cv.nohgd$lambda.1se))
  coefs[[as.character(a)]] = coef.stability(coef.1se, cv.nohgd$non.zero.cf)
}

@@

for (i in names(coef.stable)) {
  if (is.null(names(coef.stable$`0.8`))) next
  cfs[intersect(rownames(cfs), names(coef.stable[[i]])), i] = 1
}

@@

plots = list(); coefs = list()
for (a in alpha.values) {
  fit2 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
  autoplot(fit2, xvar='lambda', main='fit1, no HGD/IMC samples') + theme(legend.position='none') 
  lambda2 <- fit2$lambda

@@

  #plots[[as.character(a)]] = cv.nolgd$plot + ggtitle(paste('alpha=',a,sep=''))

  plots[[as.character(a)]] = arrangeGrob(cv.nolgd$plot+ggtitle('CV Performance'), cv.nolgd$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)

@@

for (i in names(coef.stable)) 
  cfs[intersect(rownames(cfs), names(coef.stable[[i]])), i] = 1
```

@@

@@

few.plots = list()
for (a in alpha.values) {
  fit0 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, nlambda=500, family='binomial') # all patients
  cv.back.few = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fit0$lambda, pts=subset(pts, Samplename %in% samples), splits=5, nfolds=20, a=a, fit=fit0, minR=0.1)
  #few.plots[[as.character(a)]] = cv.back.few$plot + ggtitle(paste('alpha=',a,sep=''))

  few.plots[[as.character(a)]] = arrangeGrob(cv.back.few$plot+ggtitle('CV Performance'), cv.back.few$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)
}
do.call(grid.arrange, c(few.plots, ncol=2, top=paste(nrow(subset(sum.patient.data, Status == 'P' & total.samples < m))," Progressors with < median # samples (",m,")")))

# Many
samples = c(nonps, as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'P' & total.samples >= m)$Patient], function(pt) pt$info$Samplename))))

many.plots = list()
for (a in alpha.values) {
  fit0 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, nlambda=500, family='binomial') # all patients

  cv.back.many = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fit0$lambda, pts=subset(pts, Samplename %in% samples), splits=5, nfolds=20, a=a, fit=fit0)
  #many.plots[[as.character(a)]] = cv.back.many$plot + ggtitle(paste('alpha=',a,sep=''))

  many.plots[[as.character(a)]] = arrangeGrob(cv.back.many$plot+ggtitle('CV Performance'), cv.back.many$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)
}
do.call(grid.arrange, c(many.plots, ncol=2, top=paste(nrow(subset(sum.patient.data, Status == 'P' & total.samples >= m))," Progressors with >= median # samples (",m,")")))

@@

# Leave one out

Leave out each patient (both P and NP) and run a new CV fit each time then predict the samples for the patient that was left out.  Fitted model includes HGD samples.

```{r leaveoneout, echo=T, message=F, warning=T}

info = do.call(rbind, lapply(patient.data, function(df) df$info))
pg.samp = lapply(patient.data, function(pt) {
  info = pt$info
  info$SampleSD = apply(pt$seg.vals[,info$Samplename], 2, sd)
  info$SampleMEAN =  apply(pt$seg.vals[,info$Samplename], 2, mean)
  info$Prediction = NA
  info$PID = unlist(lapply(info$Path.ID, function(x) unlist(strsplit(x, 'B'))[1]))
  return(info)
})

if (file.exists('/tmp/loo.Rdata')) {
  load('/tmp/loo.Rdata', verbose=T)
} else {

performance.at.1se = c(); coefs = list(); plots = list(); fits = list()
# Remove each patient (LOO)
for (pt in names(pg.samp)) {
  tmp.patient.data = patient.data[subset(sum.patient.data, Patient != pt)$Patient]
  samples = as.vector(unlist(sapply(tmp.patient.data, function(df) df$info$Samplename )))
  length(samples)

  # Fit generated on all samples, including HGD
  a = 1
  fitLOO <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
  cv = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fitLOO$lambda, 
                                pts=subset(pts, Samplename %in% samples), a=a, nfolds=10, splits=5, fit=fitLOO)
  plots[[pt]] = cv$plot + ggtitle(pt)
  fits[[pt]] = cv

  if ( length(cv$lambda.1se) > 0 ) {
    performance.at.1se = c(performance.at.1se, subset(cv$lambdas, lambda == cv$lambda.1se)$mean)

    coef.1se = as.data.frame(non.zero.coef(fit1, cv$lambda.1se))
    coefs[[pt]] = coef.stability(coef.1se, cv$non.zero.cf)

    pg.samp[[pt]]$Prediction = predict(fitLOO, dysplasia.df[pg.samp[[pt]]$Samplename,], s=cv$lambda.1se, type='response')[,1]
  } else {
    warning(paste("Patient", pt, "did not have a 1se"))
  }
}
save(plots, performance.at.1se, coefs, fits, file='/tmp/loo.Rdata')
}
